Visualization and analysis of probability distributions of large temporal data

### 2021-11-19 Rmd/04-gracsr.Rmd
Typos and consistent notation

The primary application of this work is data from the Smart Grid, Smart City (SGSC) project (2010–2014) available through the [Department of the Environment and Energy](https://data.gov.au/data/organization/doee). Half-hourly measurements of usage for more than 13,000 electricity smart meter customers is provided from October 2011 to March 2014. <!--Larger data sets include greater uncertainty about customer behavior due to growing variety of customers.--> Customers vary in size, location, and amenities such as solar panels, central heating, and air conditioning. The behavioral patterns differ amongst customers due to many temporal dependencies. Some customers use a dryer, while others dry their clothes on a line. Their weekly usage profile may reflect this. They may vary monthly, with some customers using more air conditioners or heaters than others, while having equivalent electrical equipment and weather circumstances. Some customers are night owls, while others are morning larks. Daily energy usage varies depending on whether customers stay home or work away from home. Age, lifestyle, family composition, building attributes, weather, availability of diverse electrical equipment, among other factors, make the task of properly segmenting customers into comparable energy behavior complex. The challenge is to be able to cluster consumers into these types of expected patterns, and other unexpected patterns, using only their energy usage history [@Ushakova2020-rl]. <!--To safeguard the customers' privacy, it is probable that such information is not accessible. Also, energy suppliers may not always update client information, such as property features, in a timely manner.--> There is a growing need to have methods that can examine the energy usage heterogeneity observed in smart meter data and what are some of the most common power consumption patterns.

@@

@Gupta2021-hakear provide a distance measure ($wpd$) for determining the significance of a cyclic granularity, and a ranking of multiple cyclic granularities. (This extends to harmonies, pairs of granularities that might interact with each other.) We define "significant" granularities as those with significant distributional differences across at least one category. The reason for subsetting granularities in this way, is that, clustering algorithms perform badly in the presence of nuisance variables. Granularities that do not have some difference between categories are likely to be nuisance variables. It should be noted that all of the time series in a collection may not have the same set of significant granularities.<!--This is the approach for managing this generate a subset ($S_c$) of significant granularities across a collection of time series:--> This is the approach for generating a subset ($S_c$) of significant granularities across a collection of time series:

@@

The shape and scale of the distribution of the measured variable (e.g. energy usage) affects distance calculations. Skewed distributions need to be symmetrized. Scales of individuals need to be standardized, because clustering is to select similar patterns, not magnitude of usage. (Organizing individuals based on magnitude can be achieved simply by sorting on a statistic like the average value across time.) <!--Observations often have a somewhat skewed time series distribution and their ranges might vary greatly. Statistical transformations are employed to bring all of them to the same range or normalize each observation. This is important because we are not interested in trivial clusters that vary in magnitudes, but rather in uncovering comparable patterns of distributional differences between categories.--> For the JS-based approaches, two data transformation techniques are recommended, normal-quantile transform (NQT) and robust scaling (RS).<!-- NQT is used in @Gupta2021-hakear prior sorting granularities.-- $wpd$, which is the foundation of wpd-based distances. XX wpd was not yet introduced--> While @Gupta2021-hakear already use NQT when computing $wpd$, it could be useful to standardize it for the selected set of significant granularities prior to computing the distances.

@@

We compute weighted pairwise distances $wpd$ [@Gupta2021-hakear] for all considered granularities for all observations. $wpd$ is designed to capture the maximum variation in the measured variable explained by an individual cyclic granularity or their interaction. It is estimated by the maximum pairwise distances between distributions across consecutive categories normalized by appropriate parameters. A higher value of $wpd$ indicates that some interesting patterns are expected, whereas a lower value would indicate otherwise.

Once we have chosen $wpd$ as a relevant feature for characterizing the distributions across one cyclic granularity, we have to decide how we combine differences between the multiple features (corresponding to multiple granularities) into a single number. The Euclidean distance between them is chosen, with the granularities acting as variables and $wpd$ representing the value under each variable. With this approach, we should expect the observations with similar $wpd$ values to be clustered together. Thus, this approach is useful for grouping observations that have a similar significance of patterns across different granularities. Similar significance does not imply a similar pattern, which is where this technique varies from JS-based distances, which detect differences in patterns across categories.

<!-- Consider a harmony table consisting of many harmonies, each of the form $(A, B)$, such that $A = \{ a_j: j = 1, 2, \dots, J\}$ and $B = \{ b_k: k = 1, 2, \dots, K\}$. Each household consists of a $J*K$ distributions one harmony. We compute the distributional difference -->
<!-- between $(A, B)$ for the $s^{th}$ household using $wpd_{{s}}(A,B)$. -->
<!-- $wpd_{{s}}(A,B)$ denotes the normalized weighted-pairwise distributional distances between $(A, B)$ and is a feature which measures distributional difference between harmonies. If we have $H_{N_C}$ harmonies in the harmony table, then for each household we have a vector of $wpd_{{s}}$ of $H_{N_C}$ elements with each element corresponding to one harmony. We aim to have pockets of customers showing similar periodic behavior by considering $wpd$ values for different harmonies and some time series features. The features should also characterize probability distributions of different household. -->

@@

Determining the number of clusters is typically a difficult task. Many metrics have been defined for choosing clusters. Most metrics for choosing the optimal number of clusters are based on comparing distances between observations within a class to those distances between observations between classes, which makes the assumption that there are some separated clusters. Some common procedures include the gap statistic [@tibshirani2001estimating], average silhouette width [@rousseeuw1987silhouettes], Dunn index [@Dunn1973-gk] and the separation index ($sindex$) [@Hennig2019-eq; @Hennig2014-ah]. These are constructed by balancing within-cluster homogeneity and between-cluster separation.

@@

In this work, we have chosen to use $sindex$. It is a very simple but effective metric. This is computed by averaging the smallest 10% of inter-cluster distances. It is relatively robust to nuisance observations. The value of $sindex$ always decreases, and sharp drops in value indicate candidates for the optimal number of clusters. The number of clusters corresponding to the value **before the drop** is the recommendation.

@@

<!-- Consider a harmony table consisting of many harmonies, each of the form $(A, B)$, such that $A = \{ a_j: j = 1, 2, \dots, J\}$ and $B = \{ b_k: k = 1, 2, \dots, K\}$. Each household consists of a $J*K$ distributions one harmony. We compute the distributional difference -->
<!-- between $(A, B)$ for the $s^{th}$ household using $wpd_{{s}}(A,B)$. -->
<!-- $wpd_{{s}}(A,B)$ denotes the normalized weighted-pairwise distributional distances between $(A, B)$ and is a feature which measures distributional difference between harmonies. If we have $H_{N_C}$ harmonies in the harmony table, then for each household we have a vector of $wpd_{{s}}$ of $H_{N_C}$ elements with each element corresponding to one harmony. We aim to have pockets of customers showing similar periodic behavior by considering $wpd$ values for different harmonies and some time series features. The features should also characterize probability distributions of different household. -->

@@

Figure \ref{fig:sindex-plot-validation} shows $sindex$ plotted against the number of clusters ($k$) for the range of mean differences (rows) under the different scenarios (columns). This can be used to determine the number of clusters for each scenario. When $sindex$ for each scenario are examined, it appears that $k = \{5, 4, 4\}$ is justified for scenarios S1, S2, and S3, respectively, given the sharp decrease in $sindex$ from that value of $k$. Thus, the number of clusters corresponds to the number of designs that were originally considered in each scenario.

@@

```{r mds-plot-validation, fig.cap = "MDS summary plots to illustrate the cluster sparation for the range of mean differences (rows) under the different scenarios (columns). It can be observed that clusters become more compact and separated for higher mean differences between categories across all scenarios. Between scenarios, separation is least prominent corresponding to Scenario (S3) where only one granularity is responsible for distinguishing the clusters."}

@@

Huge data sets present more complications for clustering. Clustering algorithms work well when there are well-separated clusters, with no nuisance variables or nuisance observations. When converting a series to granularities, many variables (each level of a granularity) are generated, possibly creating a slew of nuisance variables. Some customers may have a mix of energy use patterns, which could be considered nuisance observations located between major clusters. For this reason, we have chosen to select a small group of customers with relatively distinct and different patterns in order to illustrate the clustering more simply. Figure \ref{fig:hod-ind-group-png} shows the distribution across $hod$, $moy$ and $wnwd$ for the set of $24$ customers used to illustrate clustering. The customers are displayed in two columns of $12$ for space reasons. Each row, of each column, represents the profile of a single customer across different variables. Each customer is associated with an identifier of the form [$a$-$b$], where $a \in \{1, 2, \dots, 24\}$ represents the customer-prototype id and $b \in \{1, 2, \dots, 5\}$ indicates the label of the prototype in which a customer was placed. This is often a good approach to tackling a big analysis task, to start with a simpler task. The approach, however, is applicable to all customers.

@@

1. Choose a smaller subset of randomly selected $600$ customers with no implicit missing values for 2013.
2. Obtain $wpd$ for all cyclic granularities considered for these customers. It was found that $hod$ (hour-of-day), $moy$ (month-of-year) and $wnwd$ (weekend/weekday) are significant for most customers. We use these three granularities while clustering.
3. Remove customers whose data for an entire category of $hod$, $moy$ or $wnwd$ is empty. For example, a customer who does not have data for an entire month is excluded because their monthly behavior cannot be analyzed.
4. Remove customers whose energy consumption is 0 in all deciles. These are the clients whose consumption is likely to remain essentially flat and with no intriguing repeated patterns that we are interested in studying.

@@

It is common to filter data prior to fitting a supervised classification model using instance selection [@olvera2010review] which removes observations that might impede the model building. For clustering, this is analogous to identifying and removing nuisance observations. Prototype selection is more severe than instance selection, because only a handful of cases is selected. @Cutler1994-hm proposed a method called archetypal analysis which has inspired this approach but the procedure we have used follows @Fan2021-bq. First, dimension reduction such as t-SNE, MDS or PCA is used to project the data into a 2D space. Second, a few "anchor" customers far apart in 2D space are selected. Additional close neighbors to the anchors are selected. To check the selections relative to the full set of variables, we used a tour linked to a t-SNE layout using the R package `liminal` [@R-liminal]. This ensured that the final sample of clustered customers were also far apart in the high-dimensional space. (See the supplementary materials for further details.)<!--Few of these customers have similar distribution across $moy$ and some are similar in their $hod$ distribution. -->

@@

<!-- # ```{r hod-moy-wkndwday plots, fig.cap="The distribution across moy, hod and wkndwday for the selected designs. Few are similar in their hod pattern, while others are similar in $moy$ behavior. Some customers have distinct behavior as compared to all other customers.For example, although patterns across wkndwday do not look distinctly different for most customers, there is one household for whom weekend behavior is standing out from the rest."} -->
<!-- # ``` -->

@@

Clustering of the $24$ prototypes <!--and the data subset consisting of $353$ customers--> was conducted with all three distances, JS-NQT, JS-RS and WPD, and is summarized in Figures \ref{fig:opt-cluster-tsne-jsd}, \ref{fig:groups-4and5} and \ref{fig:summary-plot-wpd}. The t-SNE visualization suggests that there are four well-separated clusters. It is possible that because the representation is only 2D, the fifth group from the original prototype selection is distinctly different in high dimensions. The $sindex$ plots for the three methods indicate some disagreement: JS-NQT suggests $3$, JS-RS suggests $2$ or $5$ and WPD suggests $3$ or $5$. JS-RS would appear to match the original prototypes with the five cluster solution, but it actually differs. Even though the $sindex$ for JS-NQT suggests three clusters, the five cluster solution more closely matches the original prototypes. The WPD clustering provides a different grouping of the customers, and even though it disagrees with the original prototypes it is a useful grouping.<!--While plotting the clusters, RS-based scaling is applied to each customer, so that no customer dominates the shape of the summarized clusters because of magnitude. The plotting scales are not displayed since we want to emphasize comparable patterns rather than scales. The idea is that a customer in a cluster may have low total energy usage, but their behavior may be quite similar to a customer with high usage with respect to distributional pattern or significance across cyclic granularities.-->

@@

<!--Figure \ref{fig:opt-cluster-tsne-jsd} The $sindex$ is plotted against the number of clusters for clustering based on JS-based (JS-NQT) distances in Figure \ref{fig:opt-cluster-tsne-jsd}. The $sindex$ falls from around $0.56$ when $k=3$ to $0.54$ when $k$ increases. In our study, we use $k=4, 5$.--> Figure \ref{fig:groups-4and5} displays the summarized distributions across $4$ and $5$ clusters from JS-NQT clustering in (a) and (b) respectively, and helps to characterize each cluster. In the quantile plots the line represents the median, and the region shows the area between the $25^{th}$ and $75^{th}$ percentiles. The only difference between the four and five cluster solution is that A-4 divides further into B-4 and B-5. This additional division makes a clearer clustering, because it resolves the heterogeneity in $moy$ creating a group (B-5, customers 1-3) which has a winter peak in usage, and a group (B-4, customers 16-20) which has a start of the year peak in usage. B-2 (customers 4-9) and B-1 (customers 21-24) have distinctive $hod$ patterns but are both heterogeneous in $moy$ and $wnwd$. B-3 (customers 10-15) has peak usage at the end of the year, but is heterogeneous on $hod$ and $wnwd$. This clustering almost agrees with the clusters visible in the t-SNE plot.

<!-- A-4 is subdivided into B-4 and B-5, each of which has a distinct shape across $moy$ and $wnwd$. B-4 (id: 16-20) is distinguished by higher energy consumption in the first few months of the year and greater variability in weekend usage. B-5 (id: 1-3) has higher consumption in the middle of the year (winter months) and a similar weekday-weekend inter-quartile range. When $k = 4$ is used, these two groups merge to form A-4, which has a $moy$ profile of higher usage in both the beginning and middle of the year, which is not representative of the individuals. It may be worthwhile to compare Figures \ref{fig:opt-cluster-tsne-jsd} and \ref{fig:hod-ind-group-png} to see if the summarized distributions across groups accurately characterized the groupings. If it has, then the majority of the members of the group should have a similar profile.-->

<!-- Our methodology is useful for grouping similar distributions over $hod$ and $moy$ and they are placed closely for easy comparison. Few groups have mixed patterns across $hod$ and $moy$, but few have all customers in the group having a similar profile.-->

@@

```{r hod-ind-group-png, fig.cap = "The distribution of electricity demand across individual customers over three granularities $hod$, $moy$, and $wnwd$ are shown for the $24$ selected customers using quantile and box plots. They are split into batches of $12$ in (a) and (b), with each row in (a) or (b) representing a customer. The number indicates a unique customer id and a prototype id."}
```

```{r opt-cluster-tsne-jsd, fig.cap = "Clustering summaries: (a) t-SNE computed on the $24$ selected customers, and (b) separation index ($sindex$) for $2$-$10$ clusters using JS-NQT, JS-RS and WPD. Various choices in number of clusters would be recommended. Four clusters are visible in t-SNE, although it might hide a fifth cluster because dimension reduction to 2D may be insufficient to see the difference. JS-NQT suggests $3$, JS-RS suggests $2$ or $5$ and WPD suggests $3$ or $5$."}
#include_graphics("img/opt_clust_parenthesis.png")
```

@@

```{r combined-groups-js, fig.cap = "The distribution of electricity demand across clusters over $hod$ (a), $moy$ (b), and $wnwd$ (c) are shown through quantile plots. Group 1 (customers 1-3), Group 2 (customers 4-9), Group 3 (customers 10-15), Group 4 (customers 16-20), Group 5 (customers 21-24). Groups 2 and 5 appear to have a $hod$ pattern among its members (morning and evening peak), whereas groups 1, 3, and 5 appear to have a $moy$ pattern (higher usage in mid-year). With their conditional dependencies on $hod$ or $moy$ categories, $wnwd$ differences may not be discernible on their own. (c) demonstrates that the distribution of $hod$ conditional on $wnwd$ is also not different, except for the tails. This might be because most of these chosen customers do not have a $wnwd$ pattern. It is useful to compare the raw distributions with the summarized distributions to establish that the majority of individuals in the group share the same characteristics. To facilitate this comparison, the individual distributions are sorted by their JS-based groupings.", eval = FALSE}

@@

```{r groups-4and5, fig.cap = "Summary plots for four and five clusters from JS-NQT showing the distribution of electricity demand combined for all members over $hod$, $moy$, and $wnwd$. Groups A-1 (customers 21-24), A-2 (customers 4-9), and A-3 (customers 10-15) profiles correspond to Groups B-1, B-2, and B-3, respectively. Cluster A-4 splits into B-4 (customers 16-20) and B-5 (customers 1-3) to produce the five clusters, which better resolves the $moy$ distribution.", fig.height=5}

@@

```{r summary-plot-wpd, fig.cap = "Summary plots for three (a) and five (b) clusters from WPD approach showing the $wpd$ values of each customers across $hod$, $moy$, and $wnwd$ through a parallel coordinate plot. P-1 and P-2 groups correspond to Q-1 and Q-2, respectively. Cluster P-3 is subdivided into Q-3, Q-4, and Q-5. P-1 (customers 16, 18) is distinguished by high $wpd$ on $wnwd$ values. P-2 has lower $wpd$ than $moy$ and $wnwd$ for $hod$. P-3 operates in the opposite way as P-2, with larger $wpd$ for $hod$ in comparison to $moy$ and $wnwd$. For the $5$ cluster solution, this group is divided into Q3-Q5, which are distinct due to their different relative significance of $moy$ and $wnwd$.", fig.height=6}
```

@@

In summary, none of the methods captured the five original prototypes exactly. JS-NQT was almost identical, though, but WPD produced quite a different grouping. This is quite a reasonable result and illustrates both the difficulties of clustering to obtain a particularly expected grouping and the ability to learn unexpected patterns in the data. It is possible that the JS-based distances were distracted by the presence of nuisance variables, levels of the granularities that do not contribute to clustering. This would also be supported by the results of the validation study, where clustering was less effective in S2 and S3, where only some granularities had differences between levels. Clustering using WPD is expected to produce quite different results because it will group only by overall value of a granularity, not a particular pattern. A cluster summary like Figure \ref{fig:groups-4and5} is not possible because there may be different but equally interesting patterns (e.g. high evening $hod$ and high daytime $hod$) in the same cluster. Simply it provides information that across a collection of customers this specific cluster has interesting patterns in a granularity (e.g. $hod$). One would need to post-process the cluster to separate specific patterns.

>>>>

The primary application of this work is data from the Smart Grid, Smart City (SGSC) project (2010–2014) available through @smart-meter. Half-hourly measurements of usage for more than 13,000 electricity smart meter customers are provided from October 2011 to March 2014. <!--Larger data sets include greater uncertainty about customer behavior due to growing variety of customers.--> Customers vary in size, location, and amenities such as solar panels, central heating, and air conditioning. The behavioral patterns differ amongst customers due to many temporal dependencies. Some customers use a dryer, while others dry their clothes on a line. Their weekly usage profile may reflect this. They may vary monthly, with some customers using more air conditioners or heaters than others, while having equivalent electrical equipment and weather circumstances. Some customers are night owls, while others are morning larks. Daily energy usage varies depending on whether customers stay home or work away from home. Age, lifestyle, family composition, building attributes, weather, availability of diverse electrical equipment, among other factors, make the task of properly segmenting customers into comparable energy behavior complex. The challenge is to be able to cluster consumers into these types of expected patterns, and other unexpected patterns, using only their energy usage history [@Ushakova2020-rl]. <!--To safeguard the customers' privacy, it is probable that such information is not accessible. Also, energy suppliers may not always update client information, such as property features, in a timely manner.--> There is a growing need to have methods that can examine the energy usage heterogeneity observed in smart meter data and what are some of the most common power consumption patterns.

@@

@Gupta2021-hakear provide a distance measure ($\wpd$) for determining the significance of a cyclic granularity, and a ranking of multiple cyclic granularities. (This extends to harmonies, pairs of granularities that might interact with each other.) We define "significant" granularities as those with significant distributional differences across at least one category. The reason for subsetting granularities in this way, is that, clustering algorithms perform badly in the presence of nuisance variables. Granularities that do not have some difference between categories are likely to be nuisance variables. It should be noted that all of the time series in a collection may not have the same set of significant granularities.<!--This is the approach for managing this generate a subset ($S_c$) of significant granularities across a collection of time series:--> This is the approach for generating a subset ($S_c$) of significant granularities across a collection of time series:

@@

The shape and scale of the distribution of the measured variable (e.g. energy usage) affects distance calculations. Skewed distributions need to be symmetrized. Scales of individuals need to be standardized, because clustering is to select similar patterns, not magnitude of usage. (Organizing individuals based on magnitude can be achieved simply by sorting on a statistic like the average value across time.) <!--Observations often have a somewhat skewed time series distribution and their ranges might vary greatly. Statistical transformations are employed to bring all of them to the same range or normalize each observation. This is important because we are not interested in trivial clusters that vary in magnitudes, but rather in uncovering comparable patterns of distributional differences between categories.--> For the JS-based approaches, two data transformation techniques are recommended, normal-quantile transform (NQT) and robust scaling (RS).<!-- NQT is used in @Gupta2021-hakear prior sorting granularities.-- $\wpd$, which is the foundation of wpd-based distances. XX wpd was not yet introduced--> While @Gupta2021-hakear already use NQT when computing $\wpd$, it could be useful to standardize it for the selected set of significant granularities prior to computing the distances.

@@

We compute weighted pairwise distances $\wpd$ [@Gupta2021-hakear] for all considered granularities for all observations. $\wpd$ is designed to capture the maximum variation in the measured variable explained by an individual cyclic granularity or their interaction. It is estimated by the maximum pairwise distances between distributions across consecutive categories normalized by appropriate parameters. A higher value of $\wpd$ indicates that some interesting patterns are expected, whereas a lower value would indicate otherwise.

Once we have chosen $\wpd$ as a relevant feature for characterizing the distributions across one cyclic granularity, we have to decide how we combine differences between the multiple features (corresponding to multiple granularities) into a single number. The Euclidean distance between them is chosen, with the granularities acting as variables and $\wpd$ representing the value under each variable. With this approach, we should expect the observations with similar $\wpd$ values to be clustered together. Thus, this approach is useful for grouping observations that have a similar significance of patterns across different granularities. Similar significance does not imply a similar pattern, which is where this technique varies from JS-based distances, which detect differences in patterns across categories.

<!-- Consider a harmony table consisting of many harmonies, each of the form $(A, B)$, such that $A = \{ a_j: j = 1, 2, \dots, J\}$ and $B = \{ b_k: k = 1, 2, \dots, K\}$. Each household consists of a $J*K$ distributions one harmony. We compute the distributional difference -->
<!-- between $(A, B)$ for the $s^{th}$ household using $wpd_{{s}}(A,B)$. -->
<!-- $wpd_{{s}}(A,B)$ denotes the normalized weighted-pairwise distributional distances between $(A, B)$ and is a feature which measures distributional difference between harmonies. If we have $H_{N_C}$ harmonies in the harmony table, then for each household we have a vector of $wpd_{{s}}$ of $H_{N_C}$ elements with each element corresponding to one harmony. We aim to have pockets of customers showing similar periodic behavior by considering $\wpd$ values for different harmonies and some time series features. The features should also characterize probability distributions of different household. -->

@@

Determining the number of clusters is typically a difficult task. Many metrics have been defined for choosing clusters. Most metrics for choosing the optimal number of clusters are based on comparing distances between observations within a class to those distances between observations between classes, which makes the assumption that there are some separated clusters. Some common procedures include the gap statistic [@tibshirani2001estimating], average silhouette width [@rousseeuw1987silhouettes], Dunn index [@Dunn1973-gk] and the separation index (*sindex*) [@Hennig2019-eq; @Hennig2014-ah]. These are constructed by balancing within-cluster homogeneity and between-cluster separation.

@@

In this work, we have chosen to use *sindex*. It is a very simple but effective metric. This is computed by averaging the smallest 10% of inter-cluster distances. It is relatively robust to nuisance observations. The value of *sindex* always decreases, and sharp drops in value indicate candidates for the optimal number of clusters. The number of clusters corresponding to the value **before the drop** is the recommendation.

@@

<!-- Consider a harmony table consisting of many harmonies, each of the form $(A, B)$, such that $A = \{ a_j: j = 1, 2, \dots, J\}$ and $B = \{ b_k: k = 1, 2, \dots, K\}$. Each household consists of a $J*K$ distributions one harmony. We compute the distributional difference -->
<!-- between $(A, B)$ for the $s^{th}$ household using $wpd_{{s}}(A,B)$. -->
<!-- $wpd_{{s}}(A,B)$ denotes the normalized weighted-pairwise distributional distances between $(A, B)$ and is a feature which measures distributional difference between harmonies. If we have $H_{N_C}$ harmonies in the harmony table, then for each household we have a vector of $wpd_{{s}}$ of $H_{N_C}$ elements with each element corresponding to one harmony. We aim to have pockets of customers showing similar periodic behavior by considering $\wpd$ values for different harmonies and some time series features. The features should also characterize probability distributions of different household. -->

@@

Figure \ref{fig:sindex-plot-validation} shows *sindex* plotted against the number of clusters ($k$) for the range of mean differences (rows) under the different scenarios (columns). This can be used to determine the number of clusters for each scenario. When *sindex* for each scenario are examined, it appears that $k = \{5, 4, 4\}$ is justified for scenarios S1, S2, and S3, respectively, given the sharp decrease in *sindex* from that value of $k$. Thus, the number of clusters corresponds to the number of designs that were originally considered in each scenario.

@@

```{r mds-plot-validation, fig.cap = "MDS summary plots to illustrate the cluster separation for the range of mean differences (rows) under the different scenarios (columns). It can be observed that clusters become more compact and separated for higher mean differences between categories across all scenarios. Between scenarios, separation is least prominent corresponding to Scenario (S3) where only one granularity is responsible for distinguishing the clusters."}

@@

Huge data sets present more complications for clustering. Clustering algorithms work well when there are well-separated clusters, with no nuisance variables or nuisance observations. When converting a series to granularities, many variables (each level of a granularity) are generated, possibly creating a slew of nuisance variables. Some customers may have a mix of energy use patterns, which could be considered nuisance observations located between major clusters. For this reason, we have chosen to select a small group of customers with relatively distinct and different patterns in order to illustrate the clustering more simply. Figure \ref{fig:hod-ind-group-png} shows the distribution across *hod*, *moy* and *wnwd* for the set of $24$ customers used to illustrate clustering. The customers are displayed in two columns of $12$ for space reasons. Each row, of each column, represents the profile of a single customer across different variables. Each customer is associated with an identifier of the form [$a$-$b$], where $a \in \{1, 2, \dots, 24\}$ represents the customer-prototype id and $b \in \{1, 2, \dots, 5\}$ indicates the label of the prototype in which a customer was placed. This is often a good approach to tackling a big analysis task, to start with a simpler task. The approach, however, is applicable to all customers.

@@

1. Choose a smaller subset of randomly selected $600$ customers with no implicit missing values for 2013.
2. Obtain $\wpd$ for all cyclic granularities considered for these customers. It was found that *hod* (hour-of-day), *moy* (month-of-year) and *wnwd* (weekend/weekday) are significant for most customers. We use these three granularities while clustering.
3. Remove customers whose data for an entire category of *hod*, *moy* or *wnwd* is empty. For example, a customer who does not have data for an entire month is excluded because their monthly behavior cannot be analyzed.
4. Remove customers whose energy consumption is 0 in all deciles. These are the clients whose consumption is likely to remain essentially flat and with no intriguing repeated patterns that we are interested in studying.

@@

It is common to filter data prior to fitting a supervised classification model using instance selection [@olvera2010review] which removes observations that might impede the model building. For clustering, this is analogous to identifying and removing nuisance observations. Prototype selection is more severe than instance selection, because only a handful of cases is selected. @Cutler1994-hm proposed a method called archetypal analysis which has inspired this approach but the procedure we have used follows @Fan2021-bq. First, dimension reduction such as t-SNE, MDS or PCA is used to project the data into a 2D space. Second, a few "anchor" customers far apart in 2D space are selected. Additional close neighbors to the anchors are selected. To check the selections relative to the full set of variables, we used a tour linked to a t-SNE layout using the R package `liminal` [@R-liminal]. This ensured that the final sample of clustered customers were also far apart in the high-dimensional space. (See the supplementary materials for further details.)<!--Few of these customers have similar distribution across *moy* and some are similar in their *hod* distribution. -->

@@

<!-- # ```{r hod-moy-wkndwday plots, fig.cap="The distribution across moy, hod and wkndwday for the selected designs. Few are similar in their hod pattern, while others are similar in *moy* behavior. Some customers have distinct behavior as compared to all other customers.For example, although patterns across wkndwday do not look distinctly different for most customers, there is one household for whom weekend behavior is standing out from the rest."} -->
<!-- # ``` -->

@@

Clustering of the $24$ prototypes <!--and the data subset consisting of $353$ customers--> was conducted with all three distances, JS-NQT, JS-RS and WPD, and is summarized in Figures \ref{fig:opt-cluster-tsne-jsd}, \ref{fig:groups-4and5} and \ref{fig:summary-plot-wpd}. The t-SNE visualization suggests that there are four well-separated clusters. It is possible that because the representation is only 2D, the fifth group from the original prototype selection is distinctly different in high dimensions. The *sindex* plots for the three methods indicate some disagreement: JS-NQT suggests $3$, JS-RS suggests $2$ or $5$ and WPD suggests $3$ or $5$. JS-RS would appear to match the original prototypes with the five cluster solution, but it actually differs. Even though the *sindex* for JS-NQT suggests three clusters, the five cluster solution more closely matches the original prototypes. The WPD clustering provides a different grouping of the customers, and even though it disagrees with the original prototypes it is a useful grouping.<!--While plotting the clusters, RS-based scaling is applied to each customer, so that no customer dominates the shape of the summarized clusters because of magnitude. The plotting scales are not displayed since we want to emphasize comparable patterns rather than scales. The idea is that a customer in a cluster may have low total energy usage, but their behavior may be quite similar to a customer with high usage with respect to distributional pattern or significance across cyclic granularities.-->

@@

<!--Figure \ref{fig:opt-cluster-tsne-jsd} The *sindex* is plotted against the number of clusters for clustering based on JS-based (JS-NQT) distances in Figure \ref{fig:opt-cluster-tsne-jsd}. The *sindex* falls from around $0.56$ when $k=3$ to $0.54$ when $k$ increases. In our study, we use $k=4, 5$.--> Figure \ref{fig:groups-4and5} displays the summarized distributions across $4$ and $5$ clusters from JS-NQT clustering in (a) and (b) respectively, and helps to characterize each cluster. In the quantile plots the line represents the median, and the region shows the area between the $25^{th}$ and $75^{th}$ percentiles. The only difference between the four and five cluster solution is that A-4 divides further into B-4 and B-5. This additional division makes a clearer clustering, because it resolves the heterogeneity in *moy* creating a group (B-5, customers 1-3) which has a winter peak in usage, and a group (B-4, customers 16-20) which has a start of the year peak in usage. B-2 (customers 4-9) and B-1 (customers 21-24) have distinctive *hod* patterns but are both heterogeneous in *moy* and *wnwd*. B-3 (customers 10-15) has peak usage at the end of the year, but is heterogeneous on *hod* and *wnwd*. This clustering almost agrees with the clusters visible in the t-SNE plot.

<!-- A-4 is subdivided into B-4 and B-5, each of which has a distinct shape across *moy* and *wnwd*. B-4 (id: 16-20) is distinguished by higher energy consumption in the first few months of the year and greater variability in weekend usage. B-5 (id: 1-3) has higher consumption in the middle of the year (winter months) and a similar weekday-weekend inter-quartile range. When $k = 4$ is used, these two groups merge to form A-4, which has a *moy* profile of higher usage in both the beginning and middle of the year, which is not representative of the individuals. It may be worthwhile to compare Figures \ref{fig:opt-cluster-tsne-jsd} and \ref{fig:hod-ind-group-png} to see if the summarized distributions across groups accurately characterized the groupings. If it has, then the majority of the members of the group should have a similar profile.-->

<!-- Our methodology is useful for grouping similar distributions over *hod* and *moy* and they are placed closely for easy comparison. Few groups have mixed patterns across *hod* and *moy*, but few have all customers in the group having a similar profile.-->

@@

```{r hod-ind-group-png, fig.cap = "The distribution of electricity demand across individual customers over three granularities *hod*, *moy*, and *wnwd* are shown for the $24$ selected customers using quantile and box plots. They are split into batches of $12$ in (a) and (b), with each row in (a) or (b) representing a customer. The number indicates a unique customer id and a prototype id."}
```

```{r opt-cluster-tsne-jsd, fig.cap = "Clustering summaries: (a) t-SNE computed on the $24$ selected customers, and (b) separation index (*sindex*) for $2$-$10$ clusters using JS-NQT, JS-RS and WPD. Various choices in number of clusters would be recommended. Four clusters are visible in t-SNE, although it might hide a fifth cluster because dimension reduction to 2D may be insufficient to see the difference. JS-NQT suggests $3$, JS-RS suggests $2$ or $5$ and WPD suggests $3$ or $5$."}
#include_graphics("img/opt_clust_parenthesis.png")
```

@@

```{r combined-groups-js, fig.cap = "The distribution of electricity demand across clusters over *hod* (a), *moy* (b), and *wnwd* (c) are shown through quantile plots. Group 1 (customers 1-3), Group 2 (customers 4-9), Group 3 (customers 10-15), Group 4 (customers 16-20), Group 5 (customers 21-24). Groups 2 and 5 appear to have a *hod* pattern among its members (morning and evening peak), whereas groups 1, 3, and 5 appear to have a *moy* pattern (higher usage in mid-year). With their conditional dependencies on *hod* or *moy* categories, *wnwd* differences may not be discernible on their own. (c) demonstrates that the distribution of *hod* conditional on *wnwd* is also not different, except for the tails. This might be because most of these chosen customers do not have a *wnwd* pattern. It is useful to compare the raw distributions with the summarized distributions to establish that the majority of individuals in the group share the same characteristics. To facilitate this comparison, the individual distributions are sorted by their JS-based groupings.", eval = FALSE}

@@

```{r groups-4and5, fig.cap = "Summary plots for four and five clusters from JS-NQT showing the distribution of electricity demand combined for all members over *hod*, *moy*, and *wnwd*. Groups A-1 (customers 21-24), A-2 (customers 4-9), and A-3 (customers 10-15) profiles correspond to Groups B-1, B-2, and B-3, respectively. Cluster A-4 splits into B-4 (customers 16-20) and B-5 (customers 1-3) to produce the five clusters, which better resolves the *moy* distribution.", fig.height=5}

@@

(ref:summary-plot-wpd) "Summary plots for three (a) and five (b) clusters from WPD approach showing the $\wpd$ values of each customers across *hod*, *moy*, and *wnwd* through a parallel coordinate plot. P-1 and P-2 groups correspond to Q-1 and Q-2, respectively. Cluster P-3 is subdivided into Q-3, Q-4, and Q-5. P-1 (customers 16, 18) is distinguished by high $\wpd$ on *wnwd* values. P-2 has lower $\wpd$ than *moy* and *wnwd* for *hod*. P-3 operates in the opposite way as P-2, with larger $\wpd$ for *hod* in comparison to *moy* and *wnwd*. For the $5$ cluster solution, this group is divided into Q3-Q5, which are distinct due to their different relative significance of *moy* and *wnwd*."

```{r summary-plot-wpd, fig.cap = "(ref:summary-plot-wpd)", fig.height=6}
```

@@

In summary, none of the methods captured the five original prototypes exactly. JS-NQT was almost identical, though, but WPD produced quite a different grouping. This is quite a reasonable result and illustrates both the difficulties of clustering to obtain a particularly expected grouping and the ability to learn unexpected patterns in the data. It is possible that the JS-based distances were distracted by the presence of nuisance variables, levels of the granularities that do not contribute to clustering. This would also be supported by the results of the validation study, where clustering was less effective in S2 and S3, where only some granularities had differences between levels. Clustering using WPD is expected to produce quite different results because it will group only by overall value of a granularity, not a particular pattern. A cluster summary like Figure \ref{fig:groups-4and5} is not possible because there may be different but equally interesting patterns (e.g. high evening *hod* and high daytime *hod*) in the same cluster. Simply it provides information that across a collection of customers this specific cluster has interesting patterns in a granularity (e.g. *hod*). One would need to post-process the cluster to separate specific patterns.
